---
title: 'Multivariate fundamentals: Rotation (PCA)'
author: "Ronny A. Hernández Mora"
date: "11/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(readxl)
library(janitor)
library(readr)
library(lattice)

options(scipen = 99)
```

This is the lab 2 in the course renv 690 at the University of Alberta.

Videos for this lab are:

 - [Rotation, PCA](https://www.youtube.com/watch?v=LN2guRwT75k)
 - [PCA continued, matrix rotation](https://www.youtube.com/watch?v=GkHIV4mQcTs)

## Matrix rotation in R

We are going to create a matrix with 3 variables and 10 rows with 0's and 1's
values only:
```{r matrix}
data_points <- as.vector(sample(x = c(0, 1), size = 30, replace = TRUE))

dataset <- matrix(data_points, nrow = 10, ncol = 3)
```

Now, we are going to create a rotation matrix. First we need to create a formula
to transform degrees to radians. Then, we are going to use this value to
generate a rotation matrix:
```{r rotation_matrix}
# Degrees to radians
deg_to_rad <- function(deg) {
 (deg * pi) / (180)
}

# Transform 30 degrees to radians
f2 <- deg_to_rad(30)

# Create the rotation matrix
## Rotation around z-axis
rotation_matrix <- matrix(c(cos(f2), sin(f2), 0, 
                            -sin(f2), cos(f2), 0,
                            0, 0, 1 ),
                          nrow = 3, ncol = 3)
```

Finally, we are going to show how this matrix can look like in a plot with the
original data points and the rotated data points:
```{r}
rotation <- as.matrix(dataset) %*% as.matrix(rotation_matrix)

plot(dataset[ , 1], dataset[ , 2], xlim = c(-0.5, 1.5), 
     ylim = c(-0.5, 1.5), col = "blue")

points(rotation[ , 1], rotation[ , 2], col = "red")
```

## Principal component analysis

For this example, we are going to use the dataset `USArrests`
```{r}
arrests <- USArrests %>% 
  clean_names()
  
glimpse(arrests)
```

Now, we can run a PCA on this dataset
```{r}
arrests_pca <- princomp(arrests, cor = T)
```

We have created an object with the results from the PCA. Let's check the ouputs
from the model:
```{r}
arrests_pca$loadings
arrests_pca$score
summary(arrests_pca)
eigen(cor(arrests))
eigen(cor(arrests))$values/4
```

Finally we can create a biplot of our model:
```{r}
biplot(arrests_pca, choices = c(1, 2))
```

## PCA for spatial dataset

### Read the data

```{r}
alberta_climate <- read_csv("data/lab_2/ab_climate.csv") %>% 
  clean_names() %>% 
  mutate(ecosys = as.factor(ecosys))

glimpse(alberta_climate)
```

### Plot the data

```{r}
temp_colors <- colorRampPalette(c("blue", "lightblue", "yellow", "red"))(100)

levelplot(mat ~ x * y,
          data = alberta_climate,
          aspect = "iso", cuts = 99, col.regions = temp_colors)

p_colors = colorRampPalette(c("brown", "yellow", 
                          "darkgreen", "darkblue"))(100)

levelplot(log(map) ~ x * y,
          data = alberta_climate,
          aspect = "iso", cuts = 99, 
          col.regions = p_colors)

# Try with ggplot
ggplot(data = alberta_climate, aes(x = x, y = y, fill = log(map))) +
  geom_tile() +
  scale_fill_viridis_c()


levelplot(ecosys ~ x * y,
          data = alberta_climate,
          aspect = "iso", cuts = 20,
          col.regions = rainbow(21))

# Try with ggplot
ggplot(data = alberta_climate, aes(x = x, y = y, fill = ecosys)) +
  geom_tile() +
  scale_fill_viridis_d()
```

### Perform a PCA

```{r}

```



