---
title: "03-factor_analysis"
author: "Ronny A. Hernández Mora"
date: "19/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(dplyr)
library(janitor)
library(glue)
```

 - A factor analysis maximizes the concordance of eigenvectors with
 components (called factors)
 - Factors can be though as the underlying mechanisms that cause the relations
 between the variables we measure.

## Import data

 - Data for the lab is in this [link]("http://tinyurl.com/mv690/lab/2B/data")
 
```{r}
## Original lab code
# dat2 = read.csv("USArrests2.csv")
# View(dat2)
# row.names(dat2)=dat2$State
# dat2 = dat2[,-1]
# View(dat2)

## Tidyverse style
usa_arrests <- read_csv("data/lab_3/USArrests2.csv") %>% 
  clean_names()

glimpse(usa_arrests)
```

## PCA

 - Build a PCA and check variance explained
```{r}
## Original lab code
# biplot(pca_1)
# out2 = princomp(dat2, cor=T)
# biplot(out2)

## Tidyverse style
pca_1 <- usa_arrests %>% 
  select(-state) %>% 
  princomp(cor = T)

summary(pca_1)
```

 - If we calculate the variance explained through the Eigenvalues, we will
 be able to create a **scree plot**.
 - A scree plot allows us to check how many components we should look at it.
```{r}
## Original lab code
# summary(out2)
# varexpl = eigen(cor(dat2))$values / ncol(dat2)
# varexpl
# plot(varexpl, type="o")

## Tidyverse style
usa_arrests_pca <- usa_arrests %>% 
  select(-state)

var_explained <- eigen(cor(usa_arrests_pca))$values / ncol(usa_arrests_pca)

glue("The manually calculate variance for {names(usa_arrests_pca)} is: {var_explained}")
```

 - Now that we have calculated the variance explained, we can create the
 scree_plot
```{r}
## Original lab code
plot(var_explained, type = "o")

## With ggplot

```

## Factor analysis

```{r}
factor_analysis <- factanal(usa_arrests_pca, factors = 2,
                            rotation = "varimax",
                            scores = "regression")


# out3 = factanal(dat2, factors=2, rotation="varimax",
#                 scores="regression")
```

```{r}
plot(factor_analysis$score[ , 1:2], asp = 1, type = "n")
text(factor_analysis$score[ , 1:2], row.names(usa_arrests_pca))

library(vegan)

vec1=envfit(factor_analysis$score[,1:2],
            usa_arrests_pca,
            permutations = 0)

plot(vec1, col = "red")

factor_analysis$loadings
# print(out3, cutoff=0.4)
```

 - What is SS loadings?
 
```{r}
var_explained
```

```{r}
factor_analysis$uniquenesses
factor_analysis$correlation
```

### Principal components or factors as linear combinations

```{r}
factor_scores <- factor_analysis$scores

# fas = out3$scores
# View(fas)# Factor analysis scores
# View(dat2) # Original variables
```


```{r}
fac_1_lm = lm(factor_scores[, 1] ~ murder + assault +
                rape + urban_pop + income, usa_arrests_pca)
summary(fac_1_lm)

fac_2_lm = lm(factor_scores[, 2] ~ murder + assault +
                rape + urban_pop + income, usa_arrests_pca)
summary(fac_2_lm)
```

**Use model to predict new values**

```{r}
new_data_arrests <- read_csv("data/lab_3/USArrests2_2020.csv") %>% 
  clean_names()

fac_1_prediction = predict(fac_1_lm, new_data_arrests)

fac_2_prediction = predict(fac_2_lm, new_data_arrests)

```

