---
title: "Discriminant analysis and MANOVA"
author: "Ronny A. Hernández Mora"
date: "26/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(janitor)
library(dplyr)
library(readr)
library(ggplot2)
library(MASS)
library(candisc)
# This library masks the select from dplyr!!!

# Import data
## csv with iris data is the same as the one loaded in R
# iris <- read_csv("data/lab_4/iris.csv") %>% 
#   clean_names()
```

## Discriminant analysis

First, we are going to do a linear discriminant analysis

```{r}
iris_clean <- iris %>% 
  clean_names()
  
lda_iris <- lda(species ~ ., data = iris_clean)
```

Now that we have the analysis, we can obtain the scores (the new coordinate
position of points after the matrix rotation)

```{r}
scores_lda_iris <- predict(lda_iris, iris_clean)

scores_data <- scores_lda_iris[["x"]] %>% 
  as_tibble() %>% 
  bind_cols(scores_lda_iris[["class"]]) %>% 
  rename("class" = "...3")

ggplot(scores_data, aes(x = LD1, y = LD2, color = class)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_light()

# scores = predict(out1, iris)$x
# plot(scores_lda_iris[,1:2], col=rainbow(3)[as.factor(iris_clean$species)],
#      asp=1)
```

### Alternative with `candisc` package

 - For the `lm()` function, we want to explain the numeric variables as a
 function of species. To pass this formula to the function, we need to create
 first a matrix with the numeric variables or do it with a `cbind` as shown
 in the original lab code.
```{r}
numeric_variables <- iris_clean %>% 
  dplyr::select(-species) %>% 
  as.matrix()

model <- lm(numeric_variables ~ species, data = iris_clean)

## The formula before its not the same as:
# lm(sepal_length + sepal_width + petal_length + petal_width ~ species,
#    data = iris_clean)

candisc_results <- candisc(model, term = "species")

summary(candisc_results)

candisc_results[["structure"]] # This are the loadings

## Original R code lab
# x = lm(cbind(sepal_length, sepal_width, petal_length, petal_width) ~ species,
#      data = iris_clean)
# out2 = candisc(x, term = "species")
# summary(out2)
# out2$structure
# plot(out2, which = c(1, 2), scale = 8, var.col = "#777777", var.lwd = 1,
#      col = rainbow(3))
```

### Classifying new observations

Now, we are going to use new data to classify according to our model

```{r}
new_specimens <- read_csv("data/lab_4/iris_unknown.csv") %>% 
  clean_names() %>% 
  rename(sepal_length = sep_length,
         sepal_width = sep_width,
         petal_length = pet_length,
         petal_width = pet_width)
```


```{r}
# Create the model with the known data
training <- lda(species ~ ., iris_clean)

# Classify new data with 
predict(training, new_specimens)$class

# Let's check the confussion matrix
prediction <- predict(training, iris_clean)$class
confusionmatrix(iris_clean$species, prediction)

## Original lab code
# unknown = read.csv("iris_unknown.csv")
# View(unknown)
# out1 = lda(Species~., iris)
# predict(out1, unknown)$class
# pred1 = predict(out1, iris)$class
# library(biotools)
# confusionmatrix(iris$Species, pred1)
```






